Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-05-20 manuscript.Rmd
revisions up to L from diff trans

# Litter - dLAI correlations Figure 4
dl <-dLAIlitter
dl$Y <- with(dl, dLAI * 30.5/ndays)
dl$X <- with(dl, dLAIlitter.mean * 30.5/ndays)
dlup <- dl[dl$LAIchange == "increasing",]
dldown <- dl[dl$LAIchange == "decreasing",]
lmlitup <- lm(Y ~ X, data=dlup)
lmlitdown <- lm(Y ~ X, data=dldown)

@@

Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising CO~2~ and water availability of forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, *L* should increase with elevated [CO~2~]. We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe. We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear annual pattern. Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). Leaf lifespan, estimated from mean *L* and total annual litter production, was significantly longer under elevated [CO~2~] (p `r pval(LLtest$p.value)`). Our main finding that *L* was not responsive to elevated CO~2~ are consistent with other forest FACE studies, but contrast the positive response commonly predicted by ecosystem models.

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with eC~a~, largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy fluxes as it may modify or even reverse the direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites [@Norby2005]. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predict that *L* should increase under eC~a~ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g. @Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Understanding dynamics of *L* in response to water availability and its potential interactions with C~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated woodland exposed to eC~a~, the EucFACE experiment. The woodland at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [cf. @Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The site is part of a remnant patch  of native Cumberland Plain woodland [@Tozer2003]. The vegetation consists of a native woodland undisturbed for ca. 70 years and maintained in this location since the 1880’s. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). The potential evapotranspiration (PET) is ca. 1500mm (*to be refined!*), giving a ratio of P/PET of ca. 0.53 (*quite dry - details to be added*).

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots). There is one single dominant species (*Eucalyptus tereticornis* Sm.) with some scattered individuals of *E. amplifolia* Naudin. The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings was exposed to a CO~2~ mole fraction (C~a~) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots. The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode was reached (Fig. 1a).

\
Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers. Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).

@@

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor. Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).

>>>>

# Litter - dLAI correlations Figure 4
dl <- dLAIlitter
dl$Y <- with(dl, dLAI * 30.5/ndays)
dl$X <- with(dl, dLAIlitter.mean * 30.5/ndays)
dlup <- dl[dl$LAIchange == "increasing",]
dldown <- dl[dl$LAIchange == "decreasing",]
lmlitup <- lm(Y ~ X, data=dlup)
lmlitdown <- lm(Y ~ X, data=dldown)

@@

Canopy leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, *L* should increase with elevated [CO~2~]. We estimated *L* from diffuse canopy transmittance, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [CO~2~] in this timeframe. We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]` across the study period, but did not show a clear seasonal pattern. Instead, dynamics of *L* showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). Leaf lifespan, estimated from mean *L* and total annual litter production, was up to two months longer under elevated [CO~2~] (`r LLtest$estimate2` vs. `r LLtest$estimate1` yrs; p `r pval(LLtest$p.value)`). Our main finding that *L* was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of *L* commonly predicted by many ecosystem models.

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases with elevated atmospheric [CO~2~] (further denoted as eC~a~), largely in response to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. This potential change in *L* is of central importance for the net effect of eC~a~ on water, CO~2~ and energy vegetation-atmosphere fluxes as it may modify or even reverse the direct physiological responses to eC~a~ [@Tor-ngern2015; @Murray2012]. Despite its importance, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to Free Air CO~2~ Enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the five forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch; Rhinelander-FACE) and needle-leaved evergreen pine stands (DukeFACE) showed increased *L* in response to eC~a~ [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all five sites [@Norby2005]. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, consistent with a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Ellis2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predicted that *L* should increase under eC~a~ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and photo period-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g. @Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by the dynamics of leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Understanding dynamics of *L* in response to water availability and its potential interactions with eC~a~ is crucial if we are to improve phenology models in water-limited ecosystems [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated mature woodland exposed to eC~a~, the EucFACE experiment. The woodland at the EucFACE is frequently water-limited, as judged from the deficit between potential evapotranspiration (long-term mean of 1500mm; BOM) and mean annual precipitation of 800mm. The strong seasonal and interannual variability in precipitation at the site allowed us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a positive response of *L* to eC~a~ [cf. @Norby2011], either via changes in rates of leaf production or leaf retention (leaf lifespan). Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## Site description and experimental design
The EucFACE experiment is located in a remnant patch  of native Cumberland Plain woodland [@Tozer2003] on an ancient alluvial floodplain in western Sydney (Australia, 33^o^ 37'S, 150^o^ 44'E, 30 m a.s.l.). The vegetation consists of a native woodland undisturbed for ca. 70 years and maintained in this location since the 1880’s. The site is characterized by a humid temperate-subtropical transitional climate with a mean annual temperature of 17 ^o^C and mean annual precipitation (P) of 800 mm (Bureau of Meteorology, station 067105 in Richmond, NSW Australia; http://www.bom.gov.au). The potential evapotranspiration (PET) is ca. 1500mm (*to be refined!*), giving a ratio of P/PET of ca. 0.53. Over the last 30 years, monthly PET has exceeded monthly P 78% of the time, indicating very frequent water limitation.

\
At the site, the soil at the surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The vegetation consists of an open woodland (600-1000 trees ha^-1^, basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m^2^ha^-1^, n = 6 plots), with a single dominant species (*Eucalyptus tereticornis* Sm.) and some scattered individuals of *E. amplifolia* Naudin.  The understory stratum is  dominated by the grass *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted CO~2~ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings has been exposed to a CO~2~ mole fraction (C~a~) higher than ambient since September 2012, whereas the other three rings were identical to the CO~2~ treatment rings in infrastructure and instrumentation but received ambient C~a~, and are used as control plots. The target diurnal C~a~ for the rings exposed to eC~a~ was 150 $\mu$mol CO~2~ mol^-1^ above ambient C~a~ in full operation mode. However, C~a~ was increased gradually for six months at a rate of ca. 30 $\mu$mol CO~2~ mol^-1^ month^-1^ until full operation mode and the target eC~a~ was reached in February 2013 (Fig. 1a).

\
Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers (Campbell Scientific, Townsville, Australia). Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).

@@

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor. Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data; Ben Moore, unpublished data).
